---
title: "pigmeann data"
output: html_document
date: "2026-07-27"
---
# Packages
```{r}
library(ncdf4)
library(ggplot2)
```

# Extract 2021 pigmeann data
```{r}
rm(list=ls())
folder_path <- "/run/media/mmolinet/KER22/données elise/raw/PIGMeANN/daily"
year <- 2018
```

# Get times of 2021 oceanographic campaign
```{r}
# Sv_path <- paste0("/run/media/mmolinet/KER22/données elise/raw/acoustic/Acoustique_obsaus_", year, "_-100_-50dB_50ESU_10m.nc")
Sv_path <- "/run/media/mmolinet/KER22/données elise/raw/acoustic/LOCEAN_SOOP-BA_A_20180105T121559Z_MARIONDUFRESNE_FV02_EchointegrationAcoustic-18-38-70-120-200_END-20180201T103636Z_C-20260522T153853Z.nc"
ds_Sv <- nc_open(Sv_path)

lat_sv <- ncvar_get(ds_Sv, "latitude")
lon_sv <- ncvar_get(ds_Sv, "longitude")

time <- ncvar_get(ds_Sv, "time")

time_bis <- as.POSIXct(
  time * 86400,
  origin = "1950-01-01",
  tz = "UTC"
) 
dates <- as.Date(time_bis)

dates_unique <- unique(dates)

# create string matching pigmean filenames "YYYYMMDD"
dates_unique_str <- format(dates_unique, "%Y%m%d")
dates_all_str <-  format(dates, "%Y%m%d") # pour tous les jours de l'année donnée
print(dates_unique_str)
```

# colocate time from oceanographic campaign and pigmeann data

```{r}
# parse every folder and only keep ones from year of interest
files <- list.files(
  path = folder_path,
  pattern = paste0(".*(", paste(dates_unique_str, collapse = "|"), ")\\.nc$"),
  full.names = TRUE
)

# verif all dates found
n_dates_found <- length(files)
n_dates <- length(dates_unique)
print(c(n_dates, n_dates_found)) # ok 2018 (23) 2021 (45), 2022 (25); 2023 OK (25) missing
```

# Access pigmeann data
For the concentrations use: c_cond_{pigm}, and the following 2 masks
in the following order:
  - Where use_{pigm} == 0, set the pigment concentration to 0, this is based on the probability of absence of each pigment.
  - Where in_domain == 0, set the pixel to NaN (NOT 0), regardless of use_{pigm}, this is based on the mahalanobis distance to control extrapolation risk. It removes any pixel where the RRS does not resemble anything seen during the training of the algorithm. It is also based on pixels masked based on the RRS signal (optically complex waters).
  


# Colocaliser les pigments avec les stations : analyse des pigments + précise

- Par coordonnée de station 
- avec/sans voisinage temporel et spatial

```{r}
for (p in list(3, 5, 10)){
  # Initialisation dataframe
  list_pigs = c("Chla", "Per", "But", "Fuco", "Hex", "Allo", "Zea", "Chlb", "DvChla")
  pig_cond <- data.frame(
    time = time_bis,
    lat_sv = lat_sv,
    lon_sv = lon_sv,
    lat_pig = NA_real_,
    lon_pig = NA_real_
  )
  
  # Ajouter les colonnes pigments
  pig_cond[list_pigs] <- NA_real_
  
  
  length(dates_unique)
  
  for(date_i in dates_unique_str){
      # indice ESU à cette date
      ind <- which(dates_all_str == date_i)
  
      # ouvrir le fichier du jour i
      filename <- list.files(
        folder_path,
        pattern = paste0(".*(", paste(dates_unique_str, collapse = "|"), ")\\.nc$"),
        full.names = TRUE
      )
  
      
      if(length(filename)==0)
          next
  
      ds <- nc_open(filename[1])
      
      # récupérer les coordonnées PIGMeANN du jour
      lat_pig <- ds$dim$lat$vals
      lon_pig <- ds$dim$lon$vals
      
      # trouver les pixels PIGMeANN les plus proches des stations
      idx_lon <- sapply(
        lon_sv[ind],
        function(x) which.min(abs(lon_pig - x))
      )
  
      idx_lat <- sapply(
        lat_sv[ind],
        function(x) which.min(abs(lat_pig - x))
      )
  
      # enregistrer les coordonnées PIGMeANN associées
      pig_cond$lat_pig[ind] <- lat_pig[idx_lat]
      pig_cond$lon_pig[ind] <- lon_pig[idx_lon]
  
  
      for (pig in list_pigs){
  
          # lire les variables
          c_cond <- ncvar_get(
            ds,
            paste0("c_cond_", pig)
          )
          use <- ncvar_get(
            ds,
            paste0("use_", pig)
          )
  
          in_domain <- ncvar_get(
            ds,
            "in_domain"
          )
  
  
          # filtrage
          q <- quantile(
            c_cond,
            c(0.01,0.99),
            na.rm = TRUE
          )
  
          c_cond[c_cond < q[1] | c_cond > q[2]] <- NA
          c_cond[use == 0] <- 0
          c_cond[in_domain == 0] <- NA
  
          # print(c(
          #   dim(c_cond),
          #   length(lat_pig),
          #   length(lon_pig),
          #   colSums(!is.na(pig_cond[list_pigs]))
          # ))
          
          
  
          # extraction au voisinage du point
          for(j in seq_along(ind)){
  
              i <- ind[j]
  
              ilon <- idx_lon[j]
              ilat <- idx_lat[j]
  
  
              r <- (p - 1) / 2          # rayon
              
              lon_win <- max(1, ilon - r):min(nrow(c_cond), ilon + r)
              lat_win <- max(1, ilat - r):min(ncol(c_cond), ilat + r)
              
              pig_cond[i, pig] <- mean(
                  c_cond[lon_win, lat_win],
                  na.rm = TRUE
              )
          }
      }
  
      nc_close(ds)
  
  }
  saveRDS(pig_cond,file = paste0("pig_mask", p*p, "_1d_", year, ".rds"))
  idx <- which(!is.na(pig_cond$Chla))
  print(idx)
  print(unique(as.Date(pig_cond$time)[idx]))
}



```


# Affichage des distributions des concentration de pigments
```{r}


cols <- rainbow(length(list_pigs)) 
# Premier pigment 
par(mfrow=c(3,3)) 
for(pig in list_pigs){ 
  vals <- pig_cond[,pig] 
  vals <- vals[!is.na(vals)] 
  plot( density(vals), main = pig, xlab = "c_cond" ) 
} 

par(mfrow=c(1,1))



par(mfrow = c(3,3))

for (pig in list_pigs) {

  vals <- na.omit(pig_cond[[pig]])

  hist(
    vals,
    breaks = 30,
    main = pig,
    
    xlab = "c_cond"
  )
}

par(mfrow = c(1,1))
```

# affichage de la distribution des valeurs manquantes
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Proportion de valeurs manquantes par pigment
na_pig <- pig_cond %>%
  summarise(across(all_of(list_pigs), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Pigment",
    values_to = "NA_percent"
  )

# Graphique
ggplot(na_pig, aes(x = Pigment, y = NA_percent)) +
  geom_col() +
  theme_minimal() +
  labs(
    x = "Pigment",
    y = "Valeurs manquantes (%)",
    title = "Proportion de valeurs manquantes par pigment"
  ) +
  ylim(0, 100)
```
